Supplementary Information

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
USGS review changes to main document (complete)

*This draft manuscript is distributed solely for purposes of scientific peer review. Its content is deliberative and pre-decisional, so it must not be disclosed or released by reviewers. Because the manuscript has not yet been approved for publication by the USGS, it does not represent any official USGS finding or policy.*

@@

![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year. Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* P. Lawson & C. Lawson at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)

@@

```{r , eval=FALSE, echo=FALSE}
# This draw.io diagram that can be edited online here: https://app.diagrams.net/#HEcoClimLab%2FForestGEO-tree-rings%2Fmaster%2Fdoc%2Fmanuscript%2Ftables_figures%2FSummaryFigure.drawio
```
![**Figure 3 | Summary of top models for ring width (RW) as a function of climate (water and temperature variables), DBH, and calendar year**. Arrow shapes approximate reponses detailed in Figs. 3, 5, and S16. Each symbol indicates one species, and species are ordered alphabetically within each site. Overlapping arrows for the same species indicate that the response shape changed when *Year* was included in the model. For species on which the effect of *Year* was tested, the presence of only an arrow representing models without *Year* indicates that the effect was not included in the top models with *Year*. Interactive effects of climate and DBH are not shown.](tables_figures/SummaryFigure.png)

@@

![**Figure 4 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Models presented here include only climate variables as fixed effects. Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 6), relationships are plotted if included in the top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. Species authorities are given in Table S2.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Models presented here include climate variables and DBH as fixed effects. Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. Species authorities are given in Table S2.](tables_figures/DBH_responses.png)

@@

![**Figure 7 | Effect of year, when included in the best model, on BAI.**  Models presented here include climate variables, DBH, and year as fixed effects. For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. Species authorities are given in Table S2. ](tables_figures/Year_responses_BAI_only.png)

@@

The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional dendrochronological analysis methods were not designed to disentangle multiple, simultaneously acting drivers of tree growth, nor their implications for whole-forest productivity.
Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
Analyzed with respect to each ofthese drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
Applied across a wide range of forest types and species distributed globally across 10 sites, we have shown that tree species vary in the shapes of their functional responses with respect to size-related sensitivity to different climate variables.
Dissecting these species-specific long-term responses is essential to understanding the drivers of variability and directional changes in tree growth over the past century, and to predicting changes in forest composition and function in the future.

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig. `r SummaryFigure`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa*  (Kurz) Craib at HKK, *Liriodendron tulipifera* L. at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig. `r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that the approach presented here will prove valuable to understanding drivers of tree growth and forest change.
Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests worldwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs. `r SummaryFigure`, `r clim_sens`, `r interactions_fig`). 
Thus, the only way to analyze, understand, and predict global change effects on tree growth and forest productivity is by simultaneously evaluating these effects. 
The method we present and apply here allows doing so.

>>>>

@@

![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a generalized last squares (GLS) model is used to produce a combined model with the previously identified drivers, reconstructed diameter at breast height (DBH), and year. Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* P. Lawson & C. Lawson at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)

@@

```{r , eval=FALSE, echo=FALSE}
# This draw.io diagram that can be edited online here: https://app.diagrams.net/#HEcoClimLab%2FForestGEO-tree-rings%2Fmaster%2Fdoc%2Fmanuscript%2Ftables_figures%2FSummaryFigure.drawio
```
![**Figure 3 | Summary of top models for ring width (RW) as a function of climate (water and temperature variables), diameter at breast height (DBH), and calendar year**. Arrow shapes approximate responses detailed in Figs. 4, 6, and S16. Each symbol indicates one species, and species are ordered alphabetically within each site. Overlapping arrows for the same species indicate that the response shape changed when *Year* was included in the model. For species on which the effect of *Year* was tested, the presence of only an arrow representing models without *Year* indicates that the effect was not included in the top models with *Year*. Interactive effects of climate and DBH are not shown. BCNM through SC are codes for ten forested sites spanning 52 degrees latitude (Table 2); sites are ordered by descending mean annual temperature.](tables_figures/SummaryFigure.png)

@@

![**Figure 4 | Species-level responses of ring width (RW) to climwin-selected variables in precipitation and temperature variable groups for 10 sites.** Models presented here include only climate variables as fixed effects. Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. PPT = precipitation, PDF = precipitation day frequency, PET = potential evapotranspiration. For each species (color-coded as in Fig. 6), relationships are plotted if included in the top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

![**Figure 5 | Examples of climate - diameter at breast height (DBH) interactions for three tree species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. PPT = precipitation, PDF = precipitation day frequency, PET = potential evapotranspiration. Species authorities are given in Table S2.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

![**Figure 6 | Growth sensitivity to diameter at breast height (DBH): (a) ring width (RW), (b) basal area increment (BAI), (c) $\Delta$ aboveground biomass (AGB).** Models presented here include climate variables and DBH as fixed effects. Relationships for tree species in 10 sites (Table 2) are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. Species authorities are given in Table S2.](tables_figures/DBH_responses.png)

@@

![**Figure 7 | Effect of year, when included in the best model, on basal area increment (BAI).**  Models presented here include climate variables, reconstructed diameter at breast height (DBH), and year as fixed effects for 10 sites (Table 2). For each tree species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. Species authorities are given in Table S2. ](tables_figures/Year_responses_BAI_only.png)

@@

The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional dendrochronological analysis methods were not designed to disentangle multiple, simultaneously acting drivers of tree growth, nor their implications for whole-forest productivity.
Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
Analyzed with respect to each of these drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
Applied across a wide range of forest types and species distributed globally across 10 sites, we have shown that tree species vary in the shapes of their functional responses with respect to size-related sensitivity to different climate variables.
Dissecting these species-specific long-term responses is essential to understanding the drivers of variability and directional changes in tree growth over the past century, and to predicting changes in forest composition and function in the future.

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly RW initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig. `r SummaryFigure`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa*  (Kurz) Craib at HKK, *Liriodendron tulipifera* L. at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig. `r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that the approach presented here will prove valuable to understanding drivers of tree growth and forest change.
Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests worldwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs. `r SummaryFigure`, `r clim_sens`, `r interactions_fig`). 
Simultaneous evaluation of multiple drivers can improve analysis, understanding, and predictions of global change effects on tree growth and forest productivity.
The method we present and apply here allows doing so.
